Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-05-21 paper.Rnw
Mallar's suggestions plus fixes

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\todo[inline]{finish}
\end{abstract}

@@

Here we address the issue of limiting the number of input atlases by tuning our
algorithm, for segmentation of the entire hippocampus, using a multi-fold
experiment performed on a subset Alzheimer's Disease Neuroimaging Initiative
(ADNI) 1 dataset.  The tuning includes \todo{complete}. Based on the parameters
we find in our experiment, we validate our algorithm using all of the data
available in the ADNI Complete 1Yr sample and compare our segmentations to the
other segmentations that are available through the ADNI informatics portal.  To
ensure that we have not over-fit our parameters to the aging or
neurodegenerative brain, we also apply our segmentations to a dataset of normal
controls and individuals suffering from first episode psychosis. Finally, we
perform a leave-one-out validation experiment to determine if the subfields can
be accurately identified using our multi-atlas framework.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each
disease category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available. These labels have been generated using a
semiautomatic tool (SNT, Medtronic Surgical Navigation Technologies, Louisville,
CO; see Supplementary Materials).

@@

\paragraph{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this experiment we explore three fusion methods.

\subparagraph{Voxel-wise Majority Vote}
  Labels are propagated from all template library images to a target.  Each
  output voxel is given the most frequent label at that voxel location amongst
  all candidate labellings.  Ties are broken arbitrarily.

\subparagraph{Cross-correlation Weighted Majority Vote}
  An optimal combination of targets from the template library has previously been
  shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
  method, each template library image is ranked in similarity to each unlabelled 
  image by the normalized cross-correlation (CC) of image intensities after linear
  registration, over a region of interest (ROI) generously encompassing the 
  hippocampus.  Only the top ranked template library image labels are used in a
  voxel-wise majority vote. The ROI is heuristically defined as the extent of all
  atlas labels after linear registration to the template, dilated by three voxels
  \citep{Chakravarty2012}.  The number of top ranked template library image labels
  is a configurable parameter.

@@

\subparagraph{Normalised Mutual Information Weighted Majority Vote}
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between to images.

\paragraph{Evaluation method}

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as 
the gold standard to compare with. We compare both \mb labels and multi-atlas 
labels.  The segmentation accuracy reported is averaged over the ten validation
rounds for each parameter setting.

@@

In this experiment, the five $300\mu$-isotropic voxel Winterburn atlases are
used as the atlas library for \mb segmentation. LOOCV is conducted separately
for two different input datasets: the four Winterburn atlas subject T1 BRAVO
images (referred to as the {\em BRAVO} dataset), and the five
Winterburn atlas subject images resampled to 0.9mm-isotropic voxel resolution
(referred to as the {\em Resampled} dataset. Each subject in the
dataset is segmented by \mb with that subject's image excluded from the atlas
library. The template library consists of 3T T1 images (0.9mm-isotropic
voxels) of healthy subjects in addition to the images from the dataset being
evaluated.

@@

\paragraph{Evaluation method}  
\mb performance is measured with respect to straight-forward nearest-neighbour
resampling of the Winterburn atlas labels to 0.9mm-isotropic voxels (i.e. no 
label propagation). The volumes of each hippocampal subregion in the downsampled
Winterburn atlas labels are compared to the volumes obtained from the full 
resolution Winterburn atlas labels, the \mb labels on the downsampled Winterburn 
atlas images, and the \mb labels on the T1 BRAVO acquisitions.

@@

To validate that \mb algorithm works effectively with diseased brain images \mb
is applied to a dataset of Schizophrenia patient MR images. The resulting
segmentations are assessed for quality by comparison with expert manual
segmentations.

@@

The hippocampus were traced following a validated protocol developed by Dr Jens
Pruessner \todo{cite: (Pruessner et al., 2000)}. A recent update to this
protocol by Dr J Pruessner in 2006 allows to accurately and consistently
subdivide the hippocampus into three different subregions: head, body, and tail.

@@

\paragraph{Evaluation method}
The manual segmentation protocol used to segment the Winterburn atlases is
similar to, but different from, the protocol used to segment the SZFEP dataset.
Therefore, rather than evaluate using an overlap metric, such as DSC, which 
assumes a fixed segmentation protocol, we instead correlate whole hippocampus 
volume between \mb and manual segmentations.

@@

In this experiment we conducted 10 rounds of \mb and multi-atlas segmentation of
each of 69 subjects at a range of atlas and template library sizes, registration
algorithm (\ants or \animal), and three label fusion techniques.  Hippocampal
\mb-based segmentations using both \animal and \ants registration algorithm
demonstrate good overlap with manually derived gold-standards (Figure
\ref{fig:ADNI1-xval-k-mean}). Qualitatively, both \animal and \ants-based
segmentations demonstrate trend overlap accuracy that increases with the size of
atlas library and template library. Improvement in accuracy diminishes noticeably
with template libraries larger than ten images.

@@

In addition to an improvement in accuracy over multi-atlas-based segmentation,
\mb also shows a decrease in the variability of segmentation accuracy 
(Figure \todo{ADNI1-xval-variability}).  The size of template library 
necessary to reach a significant (p<0.5) decrease in variance and standard 
deviation grows with the size of atlas library used.  A template library of 19
images is sufficient to show significant decrease in variance and standard 
deviation for 3-7 atlases.

@@

ggplot(subset(stats.tests, direction > 0), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variability of segmentation accuracy within 10 rounds of validation per subject. 
  Standard deviation and variance is computed per subject for both \mb and 
  multi-atlas, and the distribution of these test statistics is compared with a 
  t-test. The p-value of this test, is shown on the y-axis (scaled logarithmically).
  Only points where \mb mean variability is lower multi-atlas are shown.}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
This experiment explores \mb segmentations of hippocampal subfields. To
achieve this, a leave-one-out validation is conducted in which lower-resolution
images ($0.9mm^3$ voxels) of each Winterburn atlas subject is segmented using 
the remaining Winterburn atlases. As a point of comparison, volumes of Winterburn 
atlases when downsampled to $0.9mm^3$ voxels are also computed.

In general, across hippocampal subregions the percent error in volume between 
\mb segmentations and the manual Winterburn atlas segmentations compares 
favourably to error when resampling the atlas segmentations (Figure 
\ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4, and Dentate subregions 
all show near or smaller percent errors. The Sibiculum and CA2/CA3 subregions 
show distinctly larger than resampling error.

@@

ggplot(melted, aes(y=percenterr,x=region,colour=measure)) +
  geom_boxplot() +
  labs(y="Percent error in volume", x="Region") +
  scale_colour_discrete("") + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Percent error in segmentation volume by hippocampus subregion.} 
  Percent error is measured against the volumes of of the unmodified Winterburn
  atlas segmentations.
  {\bf WA Resampled} are volumes of the manual segmentations of the Winterburn
  atlases after resampling to $0.9mm^3$.  
  {\bf MAGeT + WA Resampled} volumes are \mb segmentations of the Winterburn atlas
  images after resampling to $0.9mm^3$ voxels.
  {\bf MAGeT + WA BRAVO} volumes are \mb segmentations of T1 BRAVO images
  ($0.9mm^3$ voxels) acquired separately of four of the five Winterburn atlas
  subjects. 
  } 
  \label{fig:WAval-vol-boxplot}
\end{figure}

@@

\mb produces hippocampus segmentation volumes that are highly correlated with 
manual segmentation volumes (Figure \ref{fig:SZ-volumes}).

@@

Comparing total hippocampal volume, we find close agreement between manual
volumes, MAPER and MAGeT volumes across disease categories (Figure
\ref{fig:ADNI-scr-volumes-boxplot}, Figure \ref{fig:ADNI-scr-histogram}). FSL
and FreeSurfer both produce volumes in close agreement with each other, but
which are consistently larger than manually segmented volumes in each disease
category.

Additionally \mb volumes show a fixed bias towards larger volumes by an 
average $-205.2mm^3$, and a proportional bias towards producing smaller 
segmentations for larger hippocampi (Figure \ref{fig:ADNI1-src-Bland-Altman}.

@@

df = totals.complete
levels(melted) = c("FS" = ),
                   "FSL"= paste("FSL", round(cor(df$SNT, df$FSL),2)),
                   "MAPER"= paste("MAPER", round(cor(df$SNT, df$MAPER),2)),
                   "MAGeT"= paste("MAGeT", round(cor(df$SNT, df$FS),2)))

>>>>

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\todo[inline]{finish}

\end{abstract}

@@

Here we address the issue of limiting the number of input atlases by tuning our
algorithm, for segmentation of the entire hippocampus, using a multi-fold
experiment performed on a subset Alzheimer's Disease Neuroimaging Initiative
(ADNI) 1 dataset. Based on the parameters we find in our experiment, we validate
our algorithm using all of the data available in the ADNI Complete 1Yr sample
and compare our segmentations to the other segmentations that are available
through the ADNI informatics portal.  To ensure that we have not over-fit our
parameters to the aging or neurodegenerative brain, we also apply our
segmentations to a dataset of normal controls and individuals suffering from
first episode psychosis. Finally, we perform a leave-one-out validation
experiment to determine if the subfields can be accurately identified using our
multi-atlas framework.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available. These labels have been generated using the
SNT tool from Medtronic Surgical Navigation Technologies, Louisville, CO (see
Supplementary Materials for detailed discussion of the manual segmentation
process used).

@@

\paragraph{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this experiment we explore three fusion methods. 
\begin{description}
  \item[Voxel-wise Majority Vote]
  Labels are propagated from all template library images to a target.  Each
  output voxel is given the most frequent label at that voxel location amongst
  all candidate labellings.  Ties are broken arbitrarily.

  \item[Cross-correlation Weighted Majority Vote]
  An optimal combination of targets from the template library has previously been
  shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
  method, each template library image is ranked in similarity to each unlabelled 
  image by the normalized cross-correlation (CC) of image intensities after linear
  registration, over a region of interest (ROI) generously encompassing the 
  hippocampus.  Only the top ranked template library image labels are used in a
  voxel-wise majority vote. The ROI is heuristically defined as the extent of all
  atlas labels after linear registration to the template, dilated by three voxels
  \citep{Chakravarty2012}.  The number of top ranked template library image labels
  is a configurable parameter.

@@

  \item[Normalised Mutual Information Weighted Majority Vote]
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between to images.
\end{description}
\paragraph{Evaluation method}

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as 
the gold standard to compare with.  The segmentation accuracy reported is averaged over the ten validation
rounds for each parameter setting.

@@

In this experiment, the five $300\mu$-isotropic voxel Winterburn atlases are
used as the atlas library for \mb segmentation. LOOCV is conducted separately
for two different input datasets: the four Winterburn atlas subject T1 BRAVO
images (referred to as the {\em BRAVO} dataset), and the five
Winterburn atlas subject images subsampled to 0.9mm-isotropic voxel resolution
(referred to as the {\em Subsampled} dataset. Each subject in the
dataset is segmented by \mb with that subject's image excluded from the atlas
library. The template library consists of 3T T1 images (0.9mm-isotropic
voxels) of healthy subjects in addition to the images from the dataset being
evaluated.

@@

\paragraph{Evaluation method}  
\mb performance is measured with respect to straight-forward nearest-neighbour
resampling of the Winterburn atlas labels to 0.9mm-isotropic voxels (i.e. no 
label propagation). The volumes of each hippocampal subregion in the subsampled
Winterburn atlas labels are compared to the volumes obtained from the full 
resolution Winterburn atlas labels, the \mb labels on the subsampled Winterburn 
atlas images, and the \mb labels on the T1 BRAVO acquisitions.

@@

To validate that \mb algorithm works effectively in the context of other
neuropsychiatric disorders, we use the Winterburn atlases with \mb to predict
the hippocampal segmentation of dataset of Schizophrenia patient MR images. The
resulting segmentations are assessed for quality by comparison with expert
manual segmentations.

@@

The hippocampus were traced following a validated protocol developed by Dr Jens
Pruessner \todo{cite: (Pruessner et al., 2000)}. A recent update to this
protocol by Dr J Pruessner in 2006 allows to accurately and consistently
subdivide the hippocampus into three different subregions: head, body, and tail.
We use these segmentations as to validate our implementation of \mb.

@@

\paragraph{Evaluation method}
The manual segmentation protocol used to segment the Winterburn atlases has some
neuroanatomical differences in comparison to the protocol used to segment the
SZFEP dataset.  Therefore, rather than evaluate using an overlap metric, such as
DSC, which assumes a fixed segmentation protocol, we instead correlate whole
hippocampus volume between \mb and manual segmentations.

\todo{done elsewhere?}

@@

In this experiment we conducted 10 rounds of \mb and multi-atlas segmentation of
each of 69 subjects at a range of atlas and template library sizes, registration
algorithm (\ants or \animal), and three label fusion techniques.  Hippocampal
\mb-based segmentations using both \animal and \ants registration algorithm
demonstrate good overlap with SNT derived gold-standards (Figure
\ref{fig:ADNI1-xval-k-mean}). Qualitatively, both \animal and \ants-based
segmentations demonstrate trend overlap accuracy that increases with the size of
atlas library and template library. Improvement in accuracy diminishes noticeably
with template libraries larger than ten images.

@@

In addition to an improvement in accuracy over multi-atlas-based segmentation,
\mb also shows a decrease in the variability of segmentation accuracy 
(Figure \ref{ADNI1-xval-variability}).  The size of template library 
necessary to reach a significant (p<0.5) decrease in variance and standard 
deviation grows with the size of atlas library used.  A template library of 19
images is sufficient to show significant decrease in variance and standard 
deviation for 3-7 atlases.

@@

ggplot(subset(stats.tests, direction > 0), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variability of segmentation accuracy within 10 rounds of validation per subject. 
  Standard deviation and variance is computed per subject for both \mb and 
  multi-atlas, and the distribution of these test statistics is compared with a 
  t-test. The p-value of this test, is shown on the y-axis (scaled logarithmically).
  Only points where \mb mean variability is lower multi-atlas are shown.}
  \todo{note log scale}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
This experiment explores \mb segmentations of hippocampal subfields. To
achieve this, a leave-one-out validation is conducted in which lower-resolution
images ($0.9mm^3$ voxels) of each Winterburn atlas subject is segmented using 
the remaining Winterburn atlases. As a point of comparison, volumes of Winterburn 
atlases when subsampled to $0.9mm^3$ voxels are also computed.

In general, across hippocampal subregions the percent error in volume between
\mb segmentations and the manual Winterburn atlas segmentations compares
favourably to error when resampling the atlas segmentations (Figure
\ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4, and Dentate
subregions all show near or smaller percent errors. The Sibiculum and CA2/CA3
subregions show distinctly larger \todo{what?} than resampling error. 
\todo{note direction of change}

@@

ggplot(melted, aes(y=percenterr,x=region,colour=measure)) +
  geom_boxplot() +
  labs(y="Percent error in volume", x="Region") +
  scale_colour_discrete("") + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Percent error in segmentation volume by hippocampus subregion.} 
  Percent error is measured against the volumes of of the unmodified Winterburn
  atlas segmentations.
  {\bf Subsampled} are volumes of the manual segmentations of the Winterburn
  atlases after resampling to $0.9mm^3$.  
  {\bf MAGeT + WA Subsampled} volumes are \mb segmentations of the Winterburn atlas
  images after resampling to $0.9mm^3$ voxels.
  {\bf MAGeT + WA BRAVO} volumes are \mb segmentations of T1 BRAVO images
  ($0.9mm^3$ voxels) acquired separately of four of the five Winterburn atlas
  subjects.} 
  \label{fig:WAval-vol-boxplot}
\end{figure}

@@

\mb produces hippocampus segmentation volumes that are highly correlated with 
manual segmentation volumes (Figure \ref{fig:SZ-volumes}). 
\todo{p-value}

@@

Comparing total hippocampal volume, we find close agreement between manual
volumes, MAPER and MAGeT volumes across disease categories (Figure
\ref{fig:ADNI-scr-volumes-boxplot}, Figure \ref{fig:ADNI-scr-histogram}). FSL
and FreeSurfer both produce volumes in close agreement with each other, but
which are consistently larger \todo{ADD AN ESTIMATE OF THE DIFFERENCE HERE} than
manually segmented volumes in each disease category.

Additionally \mb volumes show a fixed bias towards larger volumes by an 
average $-205.2mm^3$, and a proportional bias towards producing smaller 
segmentations for larger hippocampi \todo{WHEN COMPARED TO SNT SEGMENTATIONS} (Figure \ref{fig:ADNI1-src-Bland-Altman}.

Figure \ref{fig:ADNI-scr-subfields} shows the subfield volumes for the 69 ADNI1
subjects used in Experiment 1.

@@

df = totals.complete
levels(melted) = c("FS" = paste("FSL", round(cor(df$SNT, df$FS),2)),
                   "FSL"= paste("FSL", round(cor(df$SNT, df$FSL),2)),
                   "MAPER"= paste("MAPER", round(cor(df$SNT, df$MAPER),2)),
                   "MAGeT"= paste("MAGeT", round(cor(df$SNT, df$MAGeT),2)))
